### 2014-06-11 Paper1/paper1.tex
Added stiffness derivation and what appear to be minor tweaks.

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Determining the Intent of Action Despite Disturbances}
}
% Insert Author names, affiliations and corresponding author email.
\\
Justin Horowitz$^{1}$, 
James Patton$^{1,2\ast}$
\\
\bf{1} Bioengineering, University of Illinois at Chicago, Chicago, Illinois, United States
\\
\bf{2} Sensory Motor Performance Program, Rehabilitation Institute of Chicago, Chicago, Illinois, United States
\\
$\ast$ E-mail: pattonj@uic.edu
\end{flushleft}

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
We derive a novel modeling approach for determining movement intent from a force-disturbed planar arm trajectory, examine its parameter sensitivity, and test it on human reaching data. Popular modeling approaches can be inverted in a general approach, and we show here the strengths, limitations, and underlying assumptions. We found that sensitivity to parameter inaccuracies was on the order of microns and hence much lower than the inaccuracies seen in human targeted reaching. Further, most of the sensitivity resulted from uncertainty with regard to the joint stiffness. We also tested the robustness of this approach on human point-to-point reaching movements disturbed by large force pulses. As expected, the intended trajectory showed no change from undisturbed reaching for 150 ms after the disturbance onset. Beyond this point, intent change, arm stiffening, or both usually lead to deviation from undisturbed reaching. Not all disturbed reaches showed evidence of change in intent. Unlike $\lambda$ models, the reaching intent determined by our model is physiologically plausible. This success allows real-time determination of error that can free learning enhancement and treatment methods like error augmentation and challenge point from needing to prescribe explicit task goals. In addition, extraction from recorded human motion enables direct inspection of and hypothesis testing on the contents of human motor intent. While not explicitly examined, this approach is also valid for determining intent in more complex situations such as human-machine interaction, team sports, or national security.

\section*{Introduction}
As machines interact more and more with humans, there is a growing need for a rigorous way to understand what people are trying to do, even if they become disturbed, distracted, or influenced by pathology. One can assume that a person who is familiar with a particular task will try to do the same thing each time provided all other conditions are equal. This presumes that what people do is on average what they intended to do and that their goal remains unchanged. However, if there is a disturbance, it is unclear what happens to a person's intentions because action is decoupled from intention. Challenges arise in human-machine interactions because intention and changes in intention cannot be easily discerned. Hence a comprehensive method for understanding what someone means to do (rather than what they actually do or claim to have wanted to do) as they interact within uncertain environments will prove useful in any area where a person might be thwarted from their intent such as national security, operation of machinery, athletic and music performance, rehabilitation, human augmentation, and artificial intelligence.

It is important to distinguish our use of the term intent from other groups that consider \textit{motivation}\cite{mcclelland1985motives, rawolle2013relationships}, \textit{cost}\cite{todorov2002optimal, flash1985coordination}, or \textit{goal selection decisions}\cite{ziebart2010modeling}. In typical motor control studies, subjects are motivated to complete the experiment in a timely fashion and are usually explicitly provided targets to be reached. Classification of intent is prevalent in both lower \cite{strausser2011development, hargrove2013robotic} and upper limb\cite{englehart2003robust, young2012improving} prosthetics where hybrid control algorithms select from among a set of actions (walking/standing/flexion/extension/etc). While subjects may be \textit{motivated} to complete experiments with minimal effort/cost and their \textit{goal} may be to reach a target, here we use intent to describe the course of action (i.e., the trajectory of the arm) taken in service of goals and motives and not the goals nor the motives themselves. Particularly of interest is the intended course of action when the actual movement is disturbed and hence no longer matches the intent.

@@

\subsection*{Error Metric}
Our error metric used deviation in position, where unsigned error was calculated as the mean magnitude of deviation from the nominal trajectory across each of 150 bins measuring 1 millimeter in width and spaced evenly throughout the 15centimeters reach.  Mean unsigned error (MUE) was also used to summarize the overall error in each movement for sensitivity analysis. Signed error relative to the direction of disturbance, also called perpendicular deviation, was used to measure reaching accuracy in order to elucidate direction of any corrections.

@@

\subsection*{Statistical Analysis}
The MATLAB statistics toolbox package \cite{MATLAB:2008} was used to perform comparisons among subjects and trajectory sources using the nonparametric Kruskal-Wallis test with Tukey-Kramer post-hoc for group differences. Nonparametric testing was necessary due to differences in the shapes of the force-disturbed and undisturbed hand error distributions.

@@

\subsection*{Parameter Sensitivity}
Simulated error due to direct parameter uncertainty (Figure 2) reached the order of millimeters and revealed particular sensitivity to misestimation of stiffness and changes in shoulder position from trial to trial. Indirect sensitivity, which reveals how errors from other parameters are affected by changes in a given parameter, was an order of magnitude lower than direct sensitivity and again implicated stiffness estimation inaccuracy as a primary cause of extraction uncertainty. Note that this sensitivity analysis does not provide information on whether models or parameter estimates are accurate, only how sensitive the model would be if they were inaccurate.

@@

As this is the first sound method for looking at intent following disturbance, we have only the undisturbed intention for comparison. If intent can change following disturbance, it can only do so after some period of time due to delay in processing sensory information and producing new motor commands. This represents a way to falsify the method: our determined intent should not diverge from the undisturbed intent of similar motions within this window of delay. As demonstrated in Figure 4, the method passes this test.

@@

Our simple linear scaling of Burdet et al.'s stiffness model \cite{burdet2006stability} facilitated hypothesis testing, but was inherently limiting. More accurate approaches would provide us with too many free parameters through which to determine the outcome of the eventual extraction and obtain a spurious favorable result. Our results point to the possibility of the impedance changing during motion, perhaps as a function of time. As mentioned above, while this may be perceived as a shortcoming, it was necessary and can potentially be overcome in future work.

@@

% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}
Research supported by NIH NS053606.  We thank Felix Huang, Jagan Jimmy, Christine Massie, Tejas Madhavan, and Mohi Ahmed for their review of preliminary manuscript. We also the community  of the Robotics Lab at the Rehabilitation Institute of Chicago for critical commentary during the genesis of this paper and the work leading up to it.

@@

\begin{figure}[!ht]
\begin{center}
%\includegraphics[width=4in]{fig4.eps}
\end{center}
\caption{
{\bf Comparison of subjects' perpendicular deviations across disturbed and undisturbed conditions reveals that intention is statistically indistinguishable from undisturbed reaching.} Hand deviation in the presence of disturbance (black dots, black band is mean $\pm$ simultaneous 95\% confidence interval on the mean) begin to deviate from undisturbed hand deviation (green band) beginning 95 ms after the onset of disturbance. Movement intention (red dots, band) does not significantly deviate. The hand rejoins the intention approximately 500ms after the onset of disturbance.
}
\label{fig:grouptrends}
\end{figure}

@@

\begin{table}[!ht]
\caption{
\bf{Arm Parameter Values and Sensitivities}}
\begin{tabular}{|c|c|c c|c c|}
\hline
Parameter Name &
Units &
Nominal &
SD &
$S$ &
$S_T$ \\ \hline
Upper Arm Length ($L_1$) &
m &
0.33 &
0.01 &
0 &
0.10 \\
Forearm Length ($L_2$) &
m &
0.34 &
0.01 &
0 &
0.03 \\
Upper Arm Center of Mass Ratio ($\frac{L_{m1}}{L_1}$) &
1 &
0.436 &
0.0695 &
0 &
0.02 \\
Forearm Center of Mass Ratio ($\frac{L_{m2}}{L_2}$) &
1 &
0.682 &
0.0431 &
0 &
0.02 \\
Gross Body Mass ($m_g$) &
kg &
79.36 &
3.1 &
0 &
0.02 \\
Upper Arm Mass Ratio ($\frac{m_1}{m_g}$) &
1 &
0.028 &
0.0029 &
0 &
0.01 \\
Forearm Mass Ratio ($\frac{m_2}{m_g}$) &
1 &
0.022 &
0.0025 &
0 &
0.06 \\
Upper Arm Radius of Gyration Ratio &
1 &
0.322 &
0.0161 &
0 &
0.01 \\
Forearm Radius of Gyration Ratio &
1 &
0.468 &
0.0234 &
0 &
0.01 \\
Shoulder Parallel Coordinate &
m &
-0.057 &
0.02 &
0.02 &
0.18 \\
Shoulder Perpendicular Coordinate &
m &
0.88 &
0.02 &
0 &
0.08 \\
Force Sensor Miscalibration $x\text{-axis}$ &
N &
0 &
0.231 &
0 &
0.08 \\
Force Sensor Miscalibration $y\text{-axis}$ &
N &
0 &
0.1067 &
0.03 &
0.08 \\
Force Sensor Gaussian Noise SD $x\text{-axis}$ &
N &
0 &
0.1653 &
0 &
0.01 \\
Force Sensor Gaussian Noise SD $y\text{-axis}$ &
N &
0 &
0.2869 &
0 &
0.01 \\
Torque-Invariant Impedance Misestimation Ratio &
1 &
1 &
0.15 &
0.06 &
0.31 \\
Torque-Varying Impedance Misestimation Ratio &
1 &
1 &
0.15 &
0.0 &
0.15 \\
Damping-to-Stiffness Ratio ($k_d$) &
$\text{sec}^{-1}$ &
0.0833 &
0.0125 &
0 &
0.08 \\
Reflex Impedance Scale Factor &
1 &
0.02 &
0.003 &
0 &
0.04 \\
Reflex Damping to Stiffness Ratio ($g_d$) &
$\text{sec}^{-1}$ &
2 &
0.3 &
0 &
0.04 \\ \hline
\end{tabular}
\begin{flushleft}Synthetic model parameters and their associated mean (nominal) values and standard deviations (SD) used to determine the sensitivity indices of Saltelli et al. \cite{saltelli2010variance}. Also shown are the resulting sensitivity indices. Sensitivity indices less than one-thousandth were reported as zero. Total variance was $1.3 \mu m^2$, showing very small average deviations when using this approach. First-order sensitivity, $S$, can be interpreted as fraction of variance removable by perfectly correcting a factor. Total sensitivity, $S_T$, can be interpreted as the fraction left after correcting all other factors.
\end{flushleft}
\label{tab:parameters}
\end{table}

>>>>

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{I meant to do that: Extracting the intentions of action in the face of disturbances}
}
% Insert Author names, affiliations and corresponding author email.
\\
Justin Horowitz$^{1}$, 
James Patton$^{1,2\ast}$
\\
\bf{1} Bioengineering, University of Illinois at Chicago, Chicago, Illinois, United States
\\
\bf{2} Sensory Motor Performance Program, Rehabilitation Institute of Chicago, Chicago, Illinois, United States
\\
$\ast$ E-mail: pattonj@uic.edu
\end{flushleft}

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
We derive a novel modeling approach for determining movement intent from a force-disturbed planar arm trajectory, examine its parameter sensitivity, and test it on human reaching data. Popular feedforward control models     approaches can be inverted in a general approach, and we show here the strengths, limitations, and underlying assumptions. We found that sensitivity to parameter inaccuracies was on the order of microns and hence much lower than the inaccuracies seen in human targeted reaching. Further, most of the sensitivity resulted from uncertainty with regard to the joint stiffness. We also tested the robustness of this approach on human point-to-point reaching movements disturbed by large force pulses. As expected, the intended trajectory showed no change from undisturbed reaching for 150 ms after the disturbance onset. Beyond this point, intent change, arm stiffening, or both usually lead to deviation from undisturbed reaching after an average of 300 milliseconds after the onset of the disturbance, even when the intention was near the target.  Unlike the well-known $\lambda$ models, this intent determined by considering a both feedforward (predictive) and feedback controls. Knowing the intent along with the action should allow real-time determination of error, making it possible to enhance error, and also directly inspect a neurally-impaired patient's intent for deficits. This approach may also be scaled for determining intent in complex systems of team sports, crowds, or swarms.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLoS ONE authors please skip this step. 
% Author Summary not valid for PLoS ONE submissions.   
\section*{Author Summary}
In our everyday struggles with bumpy roads and bustling crowds we often make movements in ways other than our plan. Here we identify a computational modeling method that measures actions and allows us to determine what the person meant to to do. We first demonstrate this method's good performance even when there is uncertainty in the parameters of the model. Next, we evaluate this on human reaching motions to targets, and show that while the arm is bumped by a disturbance, the intention remains straight. Approximately 300 milliseconds later, the intention changes in a corrective action that counteracts the disturbance. This type of computation can be broadly useful in canceling the effects of disturbances such as turbulence while typing or piloting, and can estimate intent in many other dynamic systems from molecules to crowds of people. It is also a new tool for understanding and treating impaired movement plans in neural injury.

\section*{Introduction}
As machines interact more and more with humans, there is a growing need for a rigorous way to understand what people are trying to do, even if they become disturbed, distracted, or influenced by pathology. One can assume that a person who is familiar with a particular task will try to do the same thing each time, provided all other conditions are equal. This presumes that what people do is on average what they intended to do and that their goal remains unchanged. However, if there is a disturbance, it is unclear what happens to a person's intentions because action is decoupled from intention. Challenges arise in human-machine interactions because intention and changes in intention cannot be easily discerned. Hence a comprehensive method for understanding what someone means to do (rather than what they actually do or claim to have wanted to do) as they interact within uncertain environments will prove useful in any area where a person might be thwarted from their intent such as national security, operation of machinery, athletic and music performance, rehabilitation, human augmentation, and artificial intelligence.

It is important to distinguish our use of the term intent from other groups that consider \textit{motivation}\cite{mcclelland1985motives, rawolle2013relationships}, \textit{cost}\cite{todorov2002optimal, flash1985coordination}, or \textit{goal selection decisions}\cite{ziebart2010modeling}. In typical motor control studies, subjects are motivated to complete the experiment in a timely fashion and are usually explicitly provided targets to be reached. Classification of intent is prevalent in both lower \cite{strausser2011development, hargrove2013robotic} and upper limb\cite{englehart2003robust, young2012improving} prosthetics where hybrid control algorithms select from among a set of actions (walking/standing/flexion/extension/etc). While subjects may be \textit{motivated} to complete experiments with minimal effort/cost and their \textit{goal} may be to reach a target, here we use intent to describe the course of action (i.e., the trajectory of the arm) taken in service of goals and motives and not the goals nor the motives themselves. Particularly of interest is the intended course of action, even when the actual movement is disturbed and hence no longer matches the intent.

@@

\subsection*{Error Metric}
Our error metric used deviation in position, where unsigned error was calculated as the mean magnitude of deviation from the nominal trajectory across each of 150 bins measuring 1 millimeter in width and spaced evenly throughout the 15centimeters reach.  Mean unsigned error (MUE) was also used to summarize the overall error in each movement for sensitivity analysis. Mean and maximum signed error relative to the direction of disturbance, also called perpendicular deviation, were used to measure reaching accuracy in order to elucidate direction of any corrections.

@@

\subsection*{Statistical Analysis}
The MATLAB statistics toolbox package \cite{MATLAB:2008} was used to perform comparisons among subjects and trajectory sources using ANOVA with Tukey-Kramer post-\textit{hoc} test for group differences.

@@

\subsection*{Parameter Sensitivity}
Simulated error due to direct parameter uncertainty (Figure 2) reached the order of millimeters and revealed particular sensitivity to misestimation of stiffness and changes in shoulder position from trial to trial. Direct sensitivity, which reveals the proportion of error that could be removed by correcting inaccuracy of a parameter, was an order of magnitude lower than total sensitivity, which reveals how much error would remain if all parameters were accurate,  and both implicated stiffness estimation inaccuracy as a primary cause of extraction uncertainty. Note that this sensitivity analysis does not provide information on whether models or parameter estimates are accurate, only how sensitive the model would be if they were inaccurate.

@@

As this is the first sound method for looking at intent following disturbance, we have only the undisturbed intention for comparison. If intent can change following disturbance, it can only do so after some period of time due to delay in processing sensory information and producing new motor commands. This represents a way to falsify the method: our determined intent should not diverge from the undisturbed intent of similar motions within this window of delay. We test this in Figure 4.

Following disturbance, we compare deviation in the hand and intended paths against the maximum deviation of undisturbed movements in order to determine at what point in time changes in hand path or intent can no longer be explained by normal variation and inaccuracy. Subjects' force-disturbed hand paths could no longer be explained ($p=0.05$) by deviation in undisturbed movement beginning within at most 135 milliseconds after the onset of disturbance. By contrast, intent did not change significantly for at least 300 milliseconds after the onset of disturbance (Table 2). This passes the test described above. While there does appear to be some systematic failure of the intent model to completely reject the disturbance, this is not significant. An overcorrection beginning at approximately 200 milliseconds appears to account for change in intention in all subjects.

@@

Our simple linear scaling of Burdet et al.'s stiffness model \cite{burdet2006stability} facilitated hypothesis testing, but was inherently limiting. More accurate approaches would provide us with many free parameters through which to determine the outcome of the eventual extraction and obtain a spurious favorable result. Our results point to the possibility of the impedance changing during motion, perhaps as a function of time. As mentioned above, while this may be perceived as a shortcoming, it was necessary and can potentially be overcome in future work.

@@

% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}
Research supported by NIH NS053606.  We thank Felix Huang, Jagan Jimmy, Christine Massie, Tejas Madhavan, and Mohi Ahmed for their review of preliminary manuscript. We also thank the community  of the Robotics Lab at the Rehabilitation Institute of Chicago for critical commentary during the genesis of this paper and the work leading up to it.

@@

\begin{figure}[!ht]
\begin{center}
%\includegraphics[width=4in]{fig4.eps}
\end{center}
\caption{
{\bf Comparison of subjects' perpendicular deviations across disturbed and undisturbed conditions reveals that intention is statistically indistinguishable from undisturbed reaching for at least 300 ms.} At most 135 milliseconds after the onset of disturbance, hand deviation in the presence of disturbance (black dots, black band is mean $\pm$ half of the critical value for simultaneous comparisons at 95\% confidence interval) can no longer be explained by undisturbed hand deviation (green band is mean maximum deviation). Movement intention (red dots, band is mean deviation) does not significantly deviate until at least 300 ms after the onset of disturbance.
}
\label{fig:grouptrends}
\end{figure}

@@

\begin{table}[!ht]
\caption{
\bf{Arm Parameter Values and Sensitivities}}
\begin{tabular}{|c|c|c c|c c|}
\hline
Parameter Name &
Units &
Nominal &
SD &
$S$ &
$S_T$ \\ \hline
Upper Arm Length ($L_1$) &
m &
0.33 &
0.01 &
0 &
0.10 \\
Forearm Length ($L_2$) &
m &
0.34 &
0.01 &
0 &
0.03 \\
Upper Arm Center of Mass Ratio ($\frac{L_{m1}}{L_1}$) &
1 &
0.436 &
0.0695 &
0 &
0.02 \\
Forearm Center of Mass Ratio ($\frac{L_{m2}}{L_2}$) &
1 &
0.682 &
0.0431 &
0 &
0.02 \\
Gross Body Mass ($m_g$) &
kg &
79.36 &
3.1 &
0 &
0.02 \\
Upper Arm Mass Ratio ($\frac{m_1}{m_g}$) &
1 &
0.028 &
0.0029 &
0 &
0.01 \\
Forearm Mass Ratio ($\frac{m_2}{m_g}$) &
1 &
0.022 &
0.0025 &
0 &
0.06 \\
Upper Arm Radius of Gyration Ratio &
1 &
0.322 &
0.0161 &
0 &
0.01 \\
Forearm Radius of Gyration Ratio &
1 &
0.468 &
0.0234 &
0 &
0.01 \\
Shoulder Parallel Coordinate &
m &
-0.057 &
0.02 &
0.02 &
0.18 \\
Shoulder Perpendicular Coordinate &
m &
0.88 &
0.02 &
0 &
0.08 \\
Force Sensor Miscalibration $x\text{-axis}$ &
N &
0 &
0.231 &
0 &
0.08 \\
Force Sensor Miscalibration $y\text{-axis}$ &
N &
0 &
0.1067 &
0.03 &
0.08 \\
Force Sensor Gaussian Noise SD $x\text{-axis}$ &
N &
0 &
0.1653 &
0 &
0.01 \\
Force Sensor Gaussian Noise SD $y\text{-axis}$ &
N &
0 &
0.2869 &
0 &
0.01 \\
Torque-Invariant Impedance Misestimation Ratio &
1 &
1 &
0.15 &
0.06 &
0.31 \\
Torque-Varying Impedance Misestimation Ratio &
1 &
1 &
0.15 &
0 &
0.15 \\
Damping-to-Stiffness Ratio ($k_d$) &
$\text{sec}^{-1}$ &
0.0833 &
0.0125 &
0 &
0.08 \\
Reflex Impedance Scale Factor &
1 &
0.02 &
0.003 &
0 &
0.04 \\
Reflex Damping to Stiffness Ratio ($g_d$) &
$\text{sec}^{-1}$ &
2 &
0.3 &
0 &
0.04 \\ \hline
\end{tabular}
\begin{flushleft}Synthetic model parameters and their associated mean (nominal) values and standard deviations (SD) used to determine the sensitivity indices of Saltelli et al. \cite{saltelli2010variance}. Also shown are the resulting sensitivity indices. Sensitivity indices less than one-thousandth were reported as zero. Total variance was $1.3 \mu m^2$, showing very small average deviations when using this approach. First-order sensitivity, $S$, can be interpreted as fraction of variance removable by perfectly correcting a factor. Total sensitivity, $S_T$, can be interpreted as the fraction left after correcting all other factors.
\end{flushleft}
\label{tab:parameters}
\end{table}

\begin{table}[!ht]
\caption{
\bf{Time Elapsed until Error in Undisturbed and Disturbed Reaching are Significantly Different}}
\begin{tabular}{|c|c|c|}
\hline
Subject Number &
Hand Change (milliseconds) &
Intent Change (milliseconds) \\ \hline
1 &
105 &
305 \\
2 &
105 &
405 \\
3 &
135 &
350 \\
4 &
105 &
300 \\ \hline
\end{tabular}
\begin{flushleft} ANOVA with Tukey-Kramer post-\textit{hoc} test ($p=0.05$) was used to simultaneously compare undisturbed hand maximum deviations to force-disturbed hand and intent deviations at 5 millisecond intervals. When  a significant difference was first detected, this was noted as the onset of change.
\end{flushleft}
\label{tab:onsets}
\end{table}
